The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-09-21 manuscript/rebuttal2.Rmd

---
title: "Response to reviewer comments, part II (The five pillars)"
author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
date: "`r Sys.Date()`"
output: word_document
code_folding: hide
fig_width: 7
fig_height: 6
bibliography: 5pillars.bib
csl: plos-computational-biology.csl
---

Dear Editor,

Again, many thanks for your consideration of our manuscript.
Below is our point-by-point response to Reviewer 3's remaining comments.

@@

### Point 2

> In my view, the paper is not suitable for "anyone who analyzes biological data." After all, the
> journal is Briefings in Bioinformatics.
> Much of the material is too technical for the stated audience.
> The authors have provided "Practical guides for implementing ‘Five Pillar’ principles."
> This may be helpful to some, but it is not realistic to assume that people will consume all of
> those materials before reading the paper or that by doing so they will be able to understand all of
> the technical details in the manuscript.
> In my view, the intended audience should be clarified in the manuscript.

RESPONSE: Indeed much of the manuscript is oriented towards gold-standard practices and this could be
out of the immediate reach for many biological data analysts, but this is not a futile exercise.
Firstly, papers such as this one set out a road map of self-education which may take years to
accomplish.
Secondly, where self-education reaches its limits, they can seek workshops and other training to
strengthen these priority areas.

Therefore, we have altered the phrase in the abstract and body to recognize the journey of
bioinformaticians in training.

ABSTRACT:
> This guide is designed for bioinformatics data analysts and bioinformaticians in training,
> but should be relevant to other domain of study.

RECOMMENDATIONS:
> The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
> principles described here could equally apply to other domains of study.

@@

### Point 10

> 10. What about when your data files are too large to fit on a personal computer? Please address in
the manuscript.

RESPONSE: In accordance with this suggestion we have added the following passage to the literate
programming section.

> While it is common to run literate programs on a PC with graphical interfaces such as RStudio,
> they can also be executed in "headless" mode on high performance computing clusters or cloud
> servers using command-line interfaces to take advantage of greater computational resources.
> Headless mode is also important for integrating literate scripts into larger workflows and for
> automated testing.

@@

RESPONSE: We appreciate the reviewer's point of view.
We have removed mention of FAIR and we refer readers to the literature on this topic.
FAIR has also been removed from the pillars infographic and Figure 5 is also gone.

### R3 Point 21

> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
There are lots of problems with FAIRness in these repositories.

RESPONSE: Possibly the reviewer hasn't seen the rebuttal letter had it's own bibliography.
Anyway, since "FAIR" is not described anymore, it simplifies the description of these databases,
which satisfy most best practices. We cite the work of Bonaretti et al 2019.
We also cite Hassenruck et al 2021, who report deficiencies in these databases, but spread the blame
between database administrators, researchers and other stakeholders.
The two important passages we have amended are provided below:

### 2023-09-21 manuscript/5pillars.Rmd
addressed rev3 comments

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) literate programming, (2) code version control and sharing,
(3) compute environment control, (4) persistent data sharing and (5) documentation.
These practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

To ensure reproducibility of bioinformatics workflows, they need to be formalized in code wherever
possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
Automated processes remove the need for manual steps, which are time-consuming and
prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
leading to problems like inadvertent conversion of gene names to dates in databases and supplementary
Excel files [@Zeeberg2004-js; @Ziemann2016-pv; @Abeysooriya2021-uz].
Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.
On the other hand, some web based tools excel at reproducibility.
The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
[@The_Galaxy_Community2022-dt;@Reich2006-na].
Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).

@@

Literate programming also works neatly with the concept of the "executable paper", the idea that
the data analysis underlying an entire study can be reproduced with one or a few commands
[@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
A typical genomics/bioinformatics study could involve one large literate script, or be broken down intosmaller scripts, where each one contains the code for generating a part of the article.
A multi-script approach would benefit from a master script that executes each component.
This further makes the job of reproducibility easier.

While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
they can also be executed in "headless" mode on high performance computing clusters or cloud
servers using command-line interfaces to take advantage of greater computational resources.
Headless mode is important for integrating literate scripts into larger workflows and for
automated testing.

@@

In bioinformatics, sharing code is becoming standard practice for reproducibility and transparency,
and is a requirement for many specialized journals [@Cadwallader2022-dq].
Code sharing appears to improve the rate of article citations [@Maitner2023-pv].
One of the most popular ways to share research code is through online software repositories
with integrated version control [@Cadwallader2022-od].
A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A repository is simply a set of files under version control that represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

`git` is typically used at the command line, however it is also incorporated into integrated development
environments commonly used in bioinformatics including Rstudio, JupyterLab
[@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].
There are also several `git` clients with graphical interfaces which better allow inspection
of code changes (e.g.: [@noauthor_undated-rh; @noauthor_undated-kj]).

>>>>

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) literate programming, (2) code version control and sharing,
(3) compute environment control, (4) persistent data sharing and (5) documentation.
These practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatics data analysts and bioinformaticians in training,
but should be relevant to other domains of study.

@@

To ensure reproducibility of bioinformatics workflows, they need to be formalized in code wherever
possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
Automated processes remove the need for manual steps, which are time-consuming and
prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or web tools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
leading to problems like inadvertent conversion of gene names to dates in databases and supplementary
Excel files [@Zeeberg2004-js; @Ziemann2016-pv; @Abeysooriya2021-uz].
Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

While web tools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.
On the other hand, some web based tools excel at reproducibility.
The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
[@The_Galaxy_Community2022-dt;@Reich2006-na].
Some web tools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).

@@

Literate programming also works neatly with the concept of the "executable paper", the idea that
the data analysis underlying an entire study can be reproduced with one or a few commands
[@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
A typical genomics/bioinformatics study could involve one large literate script, or be broken down
into smaller scripts, where each one contains the code for generating a part of the article.
A multi-script approach would benefit from a master script that executes each component.
This further makes the job of reproducibility easier.

While it is common to run literate programs on a PC with graphical interfaces such as RStudio,
they can also be executed in "headless" mode on high performance computing clusters or cloud
servers using command-line interfaces to take advantage of greater computational resources.
Headless mode is also important for integrating literate scripts into larger workflows and for
automated testing.

@@

In bioinformatics, sharing code is becoming standard practice for reproducibility and transparency,
and is a requirement for many specialized journals [@Cadwallader2022-dq].
Code sharing appears to improve the rate of article citations [@Maitner2023-pv].
One of the most popular ways to share research code is through online software repositories
with integrated version control [@Cadwallader2022-od].
A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A repository is simply a set of files under version control that represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed Version Control Systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

`git` is typically used at the command line, however it is also incorporated into integrated development
environments commonly used in bioinformatics including RStudio, JupyterLab
[@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].
There are also several `git` clients with graphical interfaces which better allow inspection
of code changes (e.g.: [@noauthor_undated-rh; @noauthor_undated-kj]).
